The triad interaction of ULK1, ATG13, and FIP200 is required for ULK complex formation and autophagy

Version 1

Abstract

Abstract

In mammals, autophagosome formation, a central event in autophagy, is initiated by the ULK complex comprising ULK1/2, FIP200, ATG13, and ATG101. However, the structural basis and mechanism of the ULK complex formation remain poorly understood. Here, we predicted the core interactions organizing the ULK complex using AlphaFold, which proposed that the intrinsically disordered region of ATG13 binds to the base of the two UBL domains in the FIP200 dimer using two phenylalanines and to the tandem MIT domain of ULK1, allowing for the 1:1:2 stoichiometry of the ULK1–ATG13–FIP200 complex. We confirmed the predicted interactions by point mutations and revealed the existence of direct triad interactions among ULK1, ATG13, and FIP200 in vitro and in cells, in which each interaction was additively important for autophagic flux. These results indicate that the ULK1–ATG13–FIP200 triadic interaction is crucial for autophagosome formation and provide a structural basis and insights into the regulation mechanism of autophagy initiation in mammals.

Introduction

Macroautophagy (hereafter autophagy) is an intracellular degradation system delivering cytoplasmic components to the lysosome via autophagosomes (Mizushima and Komatsu, 2011). Autophagy contributes to intracellular quality control by degrading defective proteins and organelles, such as polyubiquitinated proteins and damaged mitochondria. Alternatively, autophagy is induced by starvation as an adaptation mechanism. These physiological roles of autophagy are associated with various diseases, including cancer and neurodegenerative disorders (Mizushima and Levine, 2020). Therefore, understanding and controlling autophagy in clinical contexts is essential.

Autophagosome formation is performed by the evolutionally conserved autophagy-related (ATG) proteins, which form several functional units to act directly on autophagosome formation (Nakatogawa, 2020). Among them, the Atg1/ULK complex is critical in initiating autophagosome formation by organizing the formation site where all the functional units, such as ATG9 vesicles and ATG8 conjugation system, colocalize (Hama, Kurikawa et al., 2023; Kannangara, Poole et al., 2021; Kishi-Itakura, Koyama-Honda et al., 2014; Ren, Nguyen et al., 2023; Suzuki, Kirisako et al., 2001; Yamamoto, Kakuta et al., 2012; Zhou, Liu et al., 2021). In mammals, the ULK complex includes ULK1/2, ATG13, ATG101, and RB1CC1/FIP200 (Mizushima, 2010; Wong, Puente et al., 2013). Structural and functional analysis has made considerable progress in the yeast counterpart, the Atg1 complex, whose core comprises Atg1 (ULK1/2 homolog), Atg13, and Atg17 (remote homolog of FIP200) (Fujioka, Suzuki et al., 2014; Li, Chung et al., 2014; Lin and Hurley, 2016; Noda, 2024; Ragusa, Stanley et al., 2012). Crystallographic analyses have unveiled the detailed interaction mode of Atg13 with Atg1 and Atg17 and the mechanism of the Atg1 complex organization (Fujioka et al., 2014). Moreover, structure-based in vitro and in vivo functional analyses have revealed that the Atg1 complex undergoes liquid-liquid phase separation to organize the autophagosome formation site on the vacuolar membrane to initiate autophagy (Fujioka, Alam et al., 2020). However, structural and functional analyses of the mammalian ULK complex lag behind that of the yeast Atg1 complex, and structural information on the interactions comprising the core of the ULK complex has long been limited to low-resolution electron microscopy (EM) data (Shi, Yokom et al., 2020). A recent preprint reported the cryo-EM structures of the human ULK1 core complex and the supercomplex between the ULK1 core complex and the class III phosphatidylinositol 3-kinase complex 1 at 4.2–6.84 Å resolution, which discovered the unprecedented interaction between these two critical complexes working at autophagy initiation (Chen, Ren et al., 2023); however, many of the interactions that construct the ULK1 core complex remained unresolved due to the insufficient quality of the density map. Recently, calcium transients on the endoplasmic reticulum surface were proposed to trigger the liquid-liquid phase separation of FIP200 to organize the autophagosome formation site, thereby initiating autophagy (Zheng, Chen et al., 2022). However, the lack of structural information makes elucidating the molecular mechanisms underlying these events challenging.

This study predicted the structure of the ULK1–ATG13–FIP200 complex using the AlphaFold2 multimer (Evans, O’Neill et al., 2021), which unveiled the detailed interactions between ATG13 and FIP200, ATG13 and ULK1, and ULK1 and FIP200. In vitro and in vivo mutational analysis confirmed all the predicted interactions, which were demonstrated to be important for autophagy. Furthermore, we found that the FIP200–ATG13 and ULK1–ATG13 interactions partially complement each other for the autophagy initiation. These findings establish the mechanism of ULK complex formation and provide a structural basis for understanding the ULK complex phase separation and regulation mechanism of autophagy in the mammalian context.

Results

Structural prediction of the ULK1–ATG13–FIP200 complex identifies FIP200-interacting residues in ATG13

Previous studies reported that the C-terminal region of ULK1 and the C-terminal intrinsically disordered region of ATG13 bind to the N-terminal region of the FIP200 homodimer (Alers, Loffler et al., 2011; Ganley, Lam du et al., 2009; Hieke, Loffler et al., 2015; Hosokawa, Hara et al., 2009; Jung, Jun et al., 2009; Papinski and Kraft, 2016; Wallot-Hieke, Verma et al., 2018). However, structural details of the interactions forming the ULK complex remain elusive. To understand the mechanism of the ULK complex formation, we predicted the structure of the FIP200 [1–634 amino acids (aa)] dimer complexed with ATG13 (isoform c, 363–517 aa) and ULK1 (801–1,050 aa) by AlphaFold2 with the AlphaFold-Multimer option (Fig. 1A, Fig. S1). For clarity, flexible loop regions in FIP200 with low pLDDT scores were removed from Figure 1A. The overall structure is consistent with the low-resolution cryo-EM model reported previously (Shi et al., 2020), including the C-shape of the FIP200 dimer and the binding of ATG13 and ULK1 to the FIP200 dimer one by one, resulting in the 1:1:2 stoichiometry of the ULK1–ATG13–FIP200 complex. The two FIP200 molecules are named FIP200A and FIP200B, where FIP200A is the one close to ULK1. Wide interactions were predicted between the four molecules, with a buried surface area of 3,658 Å2 for FIP200A–FIP200B, 2,195 Å2 for FIP200A–ATG13, 1,966 Å2 for FIP200B–ATG13, 2,140 Å2 for ATG13–ULK1, and 1,681 Å2 for FIP200A–ULK1. No interaction was observed between FIP200B and ULK1.

Structural basis of the ATG13–FIP200 interaction.

(A) Structure of the ULK1– ATG13–FIP200 core complex predicted by AlphaFold2. Flexible loop regions in FIP200 were removed from the Figure for clarity. N and C indicate N-and C-terminal regions, respectively. (B) Close-up view of the interactions between ATG13 and FIP200. The bottom panels represent the surface model of FIP200 with the coloring based on the electrostatic potentials (blue and red indicate positive and negative potentials, respectively). (C) ITC results obtained by titration of MBP-ATG13 (363–517 aa) WT or FIP3A mutant into an FIP200 (1–634 aa) solution. (D) Effect of the ATG13 FIP3A mutation on the FIP200 interaction in vivo. atg13 KO HeLa cells stably expressing FLAG-tagged ATG13 WT or FIP3A were immunoprecipitated with an anti-FLAG antibody and detected with anti-FIP200, anti-ULK1, and anti-FLAG antibodies. (E) Relative amounts of precipitated FIP200 in (D) were calculated. Solid bars indicate the means, and dots indicate the data from three independent experiments. Differences were statistically analyzed using Tukey’s multiple comparisons test.

One ATG13 molecule binds to the FIP200 dimer through the interaction of residues 365–398 and 394–482 with FIP200B and FIP200A, respectively, resulting in a 1:2 stoichiometry of the ATG13–FIP200 complex. Previous studies using hydrogen-deuterium exchange coupled to mass spectrometry identified three regions (M1–M3) in ATG13 responsible for FIP200 binding (Shi et al., 2020). Among these sites, the side-chain of Phe377 in M1, Phe397 in M2, and Phe453 in M3, was deeply inserted into the hydrophobic pockets of FIP200 (Fig. 1B). Direct interaction between FIP200 (1–634) and Atg13 (363–517) was confirmed by isothermal titration calorimetry (ITC) with a Kd value of 6.6 μM, which was severely attenuated by alanine substitution at these three phenylalanine sites (F377A, F394A, F453A; FIP3A) (Fig. 1C, Fig. S2). Furthermore, the ATG13FIP3A mutant expressed in atg13 KO cells displayed a significantly reduced coprecipitation rate of the endogenous FIP200 (Fig. 1D, E). These results confirm that ATG13 and FIP200 interact with each other via the residues predicted by AlphaFold2 in vitro and in vivo.

ULK1 and ATG13 interact via MIT-MIM interaction similar to yeast Atg1–Atg13

Next, we focused on the interaction between ULK1 and ATG13. We previously determined the crystal structure of the yeast Atg1–Atg13 complex, which revealed that the tandem two microtubule-interacting and transport (MIT) domains in Atg1 bind to the tandem two MIT-interacting motifs (MIMs) of ATG13 (Fig. 2A, right) (Fujioka et al., 2014). The predicted structure of the ULK1–ATG13 complex was quite similar to that of the yeast Atg1–Atg13 complex and was constructed by the ULK1MIT1–ATG13MIM(C) and ULK1MIT2–ATG13MIM(N) interactions (Fig. 2A, left). According to the predicted structure, Phe470 of ATG13MIM(N) and Phe512 of ATG13MIM(C) were inserted into the hydrophobic pocket of ULK1MIT2 and ULK1MIT1, respectively (Fig. 2B). Direct interaction between ULK1 (636–1,050) and Atg13 (363–517) was confirmed by ITC with a Kd value of 0.34 μM, which was severely attenuated by alanine substitution at these two phenylalanine sites (F470A, F512A; ULK2A) (Fig. 2C, Fig. S2). In atg13 KO cells, the ULK1 expression level was significantly reduced, suggesting that the lack of ATG13 severely destabilized ULK1. The exogenous expression of ATG13WT, but not of the ATG13ULK2A mutant, rescued the endogenous ULK1 expression (Fig. 2D, E). This result suggests that the ULK2A mutation in ATG13 attenuates the ATG13–ULK1 interaction, including in vivo. These results confirm that ATG13 and ULK1 interact via the residues predicted by AlphaFold2 in vitro and in vivo.

Structural basis of the ULK1–ATG13 interaction.

(A) Structure of the ULK1– ATG13 moiety of the ULK1–ATG13–FIP200 core complex in Fig. 1 A (left) and crystal structure of the yeast Atg1–Atg13 complex (right, PDB 4P1N). (B) Close-up view of the interactions between ATG13MIM(N) and ULK1MIT2 (left) and between ATG13MIM(C) and ULK1MIT1 (right). (C) ITC results obtained by titration of MBP-ULK1 (636–1050 aa) into a solution of WT or ULK2A mutant of MBP-ATG13 (363–517 aa). Due to weak binding, the KD value for the ULK2A mutant was not accurately determined. (D) Effect of the ATG13-FIP3A mutation on endogenous ULK1 levels in vivo. WT or atg13 KO HeLa cells stably expressing FLAG-tagged ATG13 WT or ULK2A mutant were lysed, and indicated proteins were detected by immunoblotting using anti-FIP200, anti-ULK1, and anti-FLAG antibodies. (E) Relative amounts of ULK1 in (D) were normalized with β-actin and calculated. Solid bars indicate the means, and dots indicate the data from three independent experiments. Differences were statistically analyzed using Tukey’s multiple comparisons test.

Direct interaction between ULK1 and FIP200 is essential for autophagy

Although AlphaFold2 predicted the direct interaction between ULK1 and FIP200 over a reasonably large contact area (1681 Å2) (Fig. 3A), a previous study reported that ULK1 required ATG13–ATG101 for FIP200 binding in vitro (Shi et al., 2020). In our predicted structure, ULK1MIT1 and ULK1MIT2 interact directly with FIP200, with the Leu967 of ULK1MIT1 and Phe997 of ULK1MIT2 inserted on the hydrophobic groove of FIP200 (Fig. 3A, B). Although we failed to measure the affinity between ULK1 and FIP200 by ITC due to aggregation upon their mixing, an in vitro pull-down assay indicated that ULK1 directly interacted with FIP200, which was significantly attenuated by the FIP2A (L967A, F997A) mutation (Fig. 3C, D). Consistently, the ULK1FIP2A mutant lost the interaction with FIP200 while partially retaining the interaction with ATG13 in cells (Fig. 3E, F). These results confirm the direct ULK1–FIP200 interaction in vitro and in vivo. Our results establish the mode of the three independent interactions, ATG13–FIP200, ATG13–ULK1, and ULK1–FIP200, in the ULK complex.

Structural basis of the ULK1–FIP200 interaction.

(A) Structure of the ULK1– FIP200 moiety of the ULK1–ATG13–FIP200 core complex in Fig. 1 A. The right panel represents the surface model of FIP200 with coloring based on the electrostatic potentials (blue and red indicate positive and negative potentials, respectively). Dotted squares indicate the regions displayed in (B). (B) Close-up view of the interactions between ULK1MIT1 and FIP200 (left) and between ULK1MIT2 and FIP200 (right). In vitro pulldown assay between GST-ULK1 (636–1050 aa) WT or FIP2A mutant with MBP-FIP200 (1–634 aa). (D) Relative amounts of precipitated MBP-FIP200 in (C) were calculated. Solid bars indicate the means, and dots indicate the data from three independent experiments. Differences were statistically analyzed using Tukey’s multiple comparisons test. (E) Effect of the ULK1 FIP2A mutation on the FIP200 interaction in vivo. Ulk1,2 DKO MEFs stably expressing FLAG-tagged ULK1 WT or FIP2A mutant were immunoprecipitated with an anti-FLAG antibody and detected with anti-FIP200, anti-ATG13, and anti-FLAG antibodies. (F) Relative amounts of precipitated FIP200 (left) and ATG13 (right) in (E) were calculated. Solid bars indicate the means, and dots indicate the data from three independent experiments. Differences were statistically analyzed using Tukey’s multiple comparisons test. (G) Halo-LC3 processing assay of ULK1 FIP2A-expressing cells. Ulk1,2 DKO MEFs stably expressing Halo-LC3 and FLAG-tagged ULK1 WT or FIP2A mutant were labeled for 15 min with 100 nm TMR-conjugated Halo ligand and incubated in starvation medium for 1 h. Cell lysates were subjected to in-gel fluorescence detection. (H) Halo processing rate in (G). The band intensity of processed Halo and Halo-LC3 in each cell line was quantified, and the relative cleavage rate was calculated as FLAG-ULK1 WT-expressing cells as 1. Solid bars indicate the means, and dots indicate the data from three independent experiments. Data were statistically analyzed using Tukey’s multiple comparisons test. (I) Colocalization of FLAG-ULK1 WT or FIP2A mutant with FIP200. Ulk1,2 DKO MEFs stably expressing FLAG-tagged ULK1 WT or FIP2A mutant were immunostained with anti-FLAG and anti-FIP200 antibodies. Scale bar, 10 μm.

To investigate the physiological importance of the ULK1–FIP200 interaction, we assessed autophagy flux using the Halo-LC3 processing assay (Rudinskiy, Bergmann et al., 2022; Yim, Yamamoto et al., 2022). Ulk1,2 DKO mouse embryonic fibroblasts (MEFs) expressing FLAG-tagged wild-type ULK1 produce a stronger band of cleaved Halo than cells which do not express ULK1. The intensity of this cleaved Halo band corresponds to the amount of Halo-LC3 delivered to lysosomes in an autophagy-dependent manner. Ulk1,2 DKO MEFs expressing the ULK1FIP2A mutant had partially but significantly reduced processed Halo bands upon starvation compared to ULK1WT-expressing cells, confirming that the direct ULK1–FIP200 interaction is crucial, although partially, for autophagy activity (Fig. 3G, H). This partial phenotype would be due to an indirect interaction between ULK1 and FIP200 via ATG13, which was supported by the observation that the ULK1FIP2A mutant colocalized with FIP200 in cells, although less efficiently than ULK1WT (Fig. 3I).

Triadic ULK1–ATG13–FIP200 interactions are critical for autophagic flux

We compared the autophagic flux of impaired ATG13–ULK1 and ATG13–FIP200 interactions to confirm the hypothesis that indirect triadic complexes are functional for autophagy. We generated knock-in (KI) cell lines that included ATG13FIP3A, ATG13ULK2A, and ATG13FU5A(F377A, F394A, F453A, F470A, F510A; FU5A) followed by a 3xFLAG-tag in the genomic ATG13 locus of HeLa cells (Fig. 4A, B). Since overexpressed ATG13-FLAG is more than 20-fold higher than its endogenous level, the mutation significance can be evaluated at more physiological expression levels using the KI cell lines (Fig. S3). The ATG13FIP3A and ATG13ULK2A mutants displayed partial colocalization with FIP200, whereas the ATG13FU5A mutant displayed scarce colocalization with FIP200 (Fig. 4C). Next, we expressed Halo-LC3 in these KI cells and performed a Halo processing assay. Cleaved Halo was partially reduced in ATG13FIP3A and ATG13ULK2A cells compared to ATG13WT cells. In contrast, cleaved Halo in ATG13FU5A cells was almost equivalent to that in the knock-out cells (Fig. 4D, E). These results suggest that ATG13–ULK1 and ATG13–FIP200 bindings complement each other in autophagy function and that ULK1, ATG13, and FIP200 directly bind to each other to organize the robust ULK complex.

ATG13–ULK1 and ATG13–FIP200 interactions complement each other in autophagy function.

(A) Schematic representation of the CRISPR-Cas9-mediated KI strategy of ATG13 mutations with FLAG tag. The C-terminally FLAG-tagged coding sequence after the exon 14 of ATG13 with or without FIP3A, ULK2A, or FU5A mutations were knocked in exon 14 of the Homo sapiens ATG13 locus. As the KI cassette expresses NeoR under the hPGK1 promoter, clones that were successfully knocked in were selected by G418. Cas9-gRNA-targeted sites in the exon 14 of H. sapiens ATG13 locus are displayed in dark blue. The homology arm for KI is presented in magenta, and the ATG13 CDS and mutations in red and cyan, respectively. NeoR is displayed in brown. Scale bar, 0.5 kilobase pair (kb). (B) Immunoblot of ATG13-FLAG KI cell lines. WT, atg13 KO, and indicated KI HeLa cells were lysed, and indicated proteins were detected by immunoblotting using anti-FIP200, anti-ULK1, and anti-FLAG antibodies. (C) Colocalization of endogenous levels of ATG13-FLAG mutants with FIP200. Indicated KI cell lines were cultured in the starvation medium for 1 h and immunostained with anti-FLAG and anti-FIP200 antibodies. Scale bar, 10 μm. (D) Halo-LC3 processing assay of ATG13-FLAG KI cell lines. Indicated KI cell lines were labeled for 15 min with 100 nm TMR-conjugated Halo ligand and incubated in starvation medium for 1 h. Cell lysates were subjected to in-gel fluorescence detection. (E) Halo processing rate in (D). The band intensity of processed Halo and Halo-LC3 in each cell line was quantified, and the relative cleavage rate was calculated as FLAG-ULK1 WT-expressing cells as 1. Solid bars indicate the means, and dots indicate the data from three independent experiments. Data were statistically analyzed using Tukey’s multiple comparisons test. (F) Schematic depiction 0of the difference between the mammalian ULK complex and the yeast Atg1 complex. Mammalian ATG13 binds to two FIP200s within the same FIP200 dimer, contributing to the stability of one ULK1 complex. Conversely, budding yeast Atg13 binds to two Atg17s within a different Atg17 dimer, allowing for endlessly repeated Atg13-Atg17 interactions. ATG101 in the ULK1 complex and Atg31-29 in the Atg1 complex are omitted for simplicity. ATG13/Atg13 is shown in yellow, ULK1/Atg1 in magenta, and FIP200/Atg17 in green. Black lines represent interactions.

Discussion

The interaction mechanism of ULK1–ATG13–FIP200, the core of the ULK complex, has been extensively studied (Alers et al., 2011; Ganley et al., 2009; Hieke et al., 2015; Hosokawa et al., 2009; Jung et al., 2009; Papinski and Kraft, 2016; Shi et al., 2020). However, most of these studies were not based on structural information, and the knowledge of the interaction mechanism remained limited. Notably, the evidence for ULK1–FIP200 interaction has been limited to in vitro studies (Shi et al., 2020), and its physiological significance has not been characterized. In this study, AlphaFold2-based point mutational analysis provides solid evidence for the direct ULK1–ATG13, ATG13–FIP200, and ULK1–FIP200 bindings in vitro and in vivo. Additionally, the triadic interaction is complementary in the cell.

One major structural difference between the mammalian ULK complex we predicted here and the yeast Atg1 complex is that one ATG13 binds to two FIP200s within the same FIP200 dimer in mammals, whereas one Atg13 binds to two Atg17s in the distinct Atg17 dimers in yeast (Fig. 4F). The latter binding mode of Atg13 bridges Atg17 dimers to each other to form a higher-order assemblage (Yamamoto, Fujioka et al., 2016), which is considered to be the Atg1 complex’s phase separation mechanism (Fujioka et al., 2020). Conversely, ATG13 cannot bridge FIP200 dimers to each other and thus cannot induce a higher-order assemblage of the ULK complex, which necessitates another mechanism for phase separation. A recent study reported that calcium transients on the endoplasmic reticulum surface trigger FIP200 phase separation (Zheng et al., 2022); however, the detailed mechanisms are unresolved. Further studies are required to elucidate the mechanisms of phase separation in organizing the autophagosome formation sites in mammals.

In addition to Atg17, budding and fission yeasts have Atg11 as a closer homolog of FIP200 (Kim, Kamada et al., 2001; Li et al., 2014; Pan, Shao et al., 2020; Sun, Li et al., 2013; Yorimitsu and Klionsky, 2005). Atg11 and Atg17 interact with Atg1 and Atg13, respectively, whereas Atg11–Atg13 and Atg17–Atg1 interactions have not been reported. Arabidopsis thaliana, which belongs to the Archaeplastida group, a supergroup distant from Opisthokonta (including yeasts and mammals), also has Atg1, Atg13, and Atg11 (Burki, Roger et al., 2020; Li et al., 2014). The A. thaliana Atg11 interacts with Atg13 but not with Atg1 (Li et al., 2014). The triadic interaction of Atg1/ULK1, ATG13, and Atg11/Atg17/FIP200 has not been reported for any species other than mammals. This triadic interaction is likely beneficial for cells that form the complicated cellular communities that comprise the metazoans. The most likely physiological significance is the fine-tuning of tissue-specific autophagic activity. Consistent with this idea, several splicing variants of ATG13 are present in vertebrates (Alers, Wesselborg et al., 2014; Hieke et al., 2015; Jung et al., 2009). In humans, at least five splicing variants are known, including the isoform 3 lacking the C-terminal ULK1-binding site (MIMs, Fig. 2A) (Alers et al., 2014; Jung et al., 2009). The avian Atg13 has seven splicing variants, some lacking parts of the FIP200 interaction site (Alers et al., 2014; Hieke et al., 2015). This study found that a single disruption of the FIP200–ATG13 or ULK1–ATG13 interaction results in a partial reduction of autophagy activity (Fig. 4D). In the individual context, the tissue-and organ-specific fine-tuning of the autophagy activity may occur by regulating the expression amount and ratio of splicing variants that lack some of the triad interactions. Generating KI mice with the point mutations identified in this study might help investigate the significance of the triadic binding at the individual level.

Materials and Methods

Cell culture

HeLa cells and MEFs were cultured in Dulbecco’s Modified Eagle medium (DMEM) (043-30085; Fujifilm Wako Pure Chemical Corporation) supplemented with 10% fetal bovine serum (FBS) (172012; Sigma-Aldrich) in a 5% CO2 incubator at 37°C. Cells were washed twice with phosphate-buffered saline (PBS) and incubated in amino acid-free DMEM (048-33575; Fujifilm Wako Pure Chemical Corporation) without FBS for starvation. atg13 KO HeLa cells (Hama et al., 2023) and Ulk1 Ulk2 double knockout MEFs (Cheong, Lindsten et al., 2011) were described previously.

Plasmids for mammalian cells

Plasmids for stable expression were generated as follows: DNA fragments encoding human ULK1 (NM_003565.4) and ATG13 (Hosokawa et al., 2009) were inserted into the retroviral plasmid pMRX-IP (puromycin-resistant marker) (Kitamura, Koshino et al., 2003; Saitoh, Nakayama et al., 2003) with a 3 × FLAG epitope tag. Site-directed mutagenesis was used to introduce ATG13 FIP3A, ATG13 ULK2A, ATG13 FU5A, and ULK1 FIP2A mutations. pMRXIB-ratLC3B was described previously (Yim et al., 2022). Plasmids for ATG13 KI were generated as follows: sgRNAs targeting the exon 14 of ATG13 (5′-GGCCTCCCCTCACGATGTCT-3′) were inserted into the BpiI site of PX458 [pSpCas9(BB)-2A-GFP; Addgene #48138]. Donor plasmids for ATG13 KI were generated as follows: 500 base-pair (bp) and 682 bp homology arms were amplified from the genomic ATG13 locus around exon 14 and inserted to flank the KI cassettes (described in Fig. 4A) in the donor plasmid (pKnockIn). The coding sequence of 3xFLAG-tagged ATG13 with or without FIP3A, ULK2A, or FU5A mutations was amplified from stable expression plasmids (described above) and inserted just behind the 5’ homology arm.

Stable expression in HeLa cells by retroviral infection

HEK293T cells were transfected with the retroviral plasmid with pCG-gag-pol and pCG-VSV-G (a gift from Dr. T. Yasui, National Institutes of Biomedical Innovation, Health and Nutrition) using FuGENE HD (E2311; Promega) for 4 h in Opti-MEM (31985-070; Gibco). After cell cultivation for 2 days in DMEM, the retrovirus-containing medium was harvested, filtered through a 0.45-µm filter unit (Ultrafree-MC; Millipore), and added to HeLa cells or MEFs with 8 μg/ml polybrene (H9268; Sigma-Aldrich). After 24 h, 2 μg/ml puromycin (P8833; Sigma-Aldrich) or 2.5 μg/ml blasticidin (022-18713; Fujifilm Wako Pure Chemical Corporation) were used to select the stable transformants.

Plasmids for protein preparation

To construct the expression plasmid encoding an N-terminal maltose binding protein (MBP) followed by an HRV3C protease site, the genes were amplified by PCR and cloned into the pET15b vector. The genes were amplified by PCR and cloned into the pET15b-MBP vector for the plasmids encoding N-terminal MBP-tagged ULK1, ATG13, and FIP200. Similarly, for the plasmids encoding N-terminal glutathione S-transferase (GST)-tagged ULK1, the genes were amplified by PCR and cloned into the pGEX6p-1 vector. The NEBuilder HiFi DNA Assembly Master Mix (New England BioLabs) was used to assemble the PCR fragments. PCR-mediated site-directed mutagenesis was used to introduce the mutations that led to the specified amino acid substitutions. All constructs were sequenced to confirm their identities.

Protein expression and purification

E. coli BL21 (DE3) cells were used to express all recombinant proteins. After cell lysis, MBP-FIP200 (1–634) was purified by affinity chromatography using Amylose Resin High Flow (New England Biolabs). Next, MBP was cleaved with the human rhinovirus 3C protease. The eluates were then desalted with 20 mM Tris-HCl, pH 8.0, and 150 mM NaCl utilizing a Bio-Scale Mini Bio-Gel P-6 desalting column (Bio-Rad Laboratories). Subsequently, the cleaved MBP was removed by reappliying FIP200 (1–634) to an Amylose Resin High Flow column. For the pull-down assay, MBP-FIP200 (1–634) eluted from the amylose resin continued to be purified on a HiLoad 26/60 Superdex 200 PG column eluted with 20 mM Tris-HCl, pH 8.0, and 150 mM NaCl.

Affinity chromatography with Amylose Resin High Flow (New England Biolabs) was used to purify the MBP-Atg13 (363–517) and MBP-ULK1 (636–1,050) proteins. They were further purified on a HiLoad 26/60 Superdex 200 PG column and eluted with a buffer containing 20 mM Tris-HCl, pH 8.0, and 150 mM NaCl. Similarly, GST-ULK1 (636–1050) underwent initial purification using a GST-accept resin (Nacalai Tesque, 09277-14). This step was followed by further purification on a HiLoad 26/60 Superdex 200 PG column using the same elution buffer as above.

In vitro pull-down assay

Purified proteins were incubated with GST-accept beads (Nacalai Tesque) at 4°C for 30 min. The beads were washed three times with PBS, and proteins were eluted with 10 mM glutathione in 50 mM Tris-HCl (pH 8.0). SDS-PAGE was used to separate the samples, and protein bands were detected by One Step CBB (BIO CRAFT).

Isothermal titration calorimetry

The binding of FIP200 (1–634) to MBP-ATG13 (363–517) and MBP-ATG13 (363–517) to MBP-ULK1 (636–1,050) was measured by ITC, with a MICROCAL PEAQ-ITC calorimeter (Malvern) and stirring at 750 rpm at 25°C. All proteins were prepared in a solution of 20 mM Tris-HCl, pH 8.0, and 150 mM NaCl. The titration of MBP-ATG13 (363–517) with FIP200 (1– 634) involved 18 injections of 2 μl of the MBP-ATG13 (363–517) solution (229 μM) at 150 s intervals into a sample cell containing 300 μl of FIP200 (1–634) (4 μM). The titration of MBP-ULK1 (636–1,050) with MBP-ATG13 (363–517) involved 18 injections of 2 μl of the MBP-ULK1 (636–1,050) solution (232 μM) at 150 s intervals into a sample cell containing 300 μl of MBP-ATG13 (363–517) (22.9 μM). The isotherm was integrated and fitted with the one-side-binding model of the Malvern MicroCal PEAQ-ITC analysis software. The error of each parameter indicates the fitting error.

Generation of KI cell lines

HeLa cells were transfected with the PX458-based plasmid expressing sgRNA for the ATG13 gene and donor plasmids using FuGENE HD (E2311; Promega). One day after transfection, the KI cells were selected with 1 mg/mL G418 for 14 days. After single clones were isolated, clones positive for the FLAG tag and negative for wild-type ATG13 were screened by immunoblotting.

Preparation of whole cell lysates and immunoblotting

HeLa cells were lysed with 0.2% n-octyl-β-D-dodecyl maltoside in lysis buffer [50 mM Tris-HCl (pH 7.5), 150 mM NaCl, 1 mM EDTA, EDTA-free protease inhibitor cocktail (04080; Nacalai Tesque)] for 15 min on ice. After centrifugation at 20,000 × g for 15 min at 4°C, the supernatants were mixed with 20% volumes of 6 × SDS-PAGE sample buffer. The samples were subjected to SDS-PAGE and transferred to Immobilon-P PVDF membranes (IPVH00010; EMD Millipore). The primary antibodies used for immunoblotting were rabbit polyclonal antibodies against FIP200 (17250-1-AP; ProteinTech), ATG13 (13273; Cell Signaling), and ULK1 (8054; Cell Signaling) and mouse monoclonal antibody against β-actin (A2228; Sigma-Aldrich). Horseradish peroxidase (HRP)-conjugated anti-rabbit IgG (111-035-144; Jackson ImmunoResearch) was used as a secondary antibody. The HRP-conjugated mouse monoclonal antibody against DDDDK-tag (M185-7; MBL) was used for FLAG tag detection. Immobilon Western Chemiluminescent HRP Substrate (P90715; EMD Millipore) was used to visualize the signals detected by an image analyzer (ImageQuant LAS 4000; Cytiva). Fiji software (ImageJ; National Institutes of Health) was used to adjust contrast and brightness and quantify the signals (Schindelin, Arganda-Carreras et al., 2012).

Immunoprecipitation

Cells were lysed with 1% 3-[(3-cholamidopropyl)-dimethylammonium]-1-propanesulfonate (CHAPS) in lysis buffer and centrifuged at 20,000 g for 15 min. The supernatants were incubated with anti-FLAG M2 affinity gel (A2220; Sigma-Aldrich) for 1 h at 4°C with gentle rotation. The beads were washed three times in washing buffer (20 mM Tris-HCl, pH 8.0, 150 mM NaCl, 1 mM EDTA, and 0.5% CHAPS), and the proteins were eluted with the SDS-PAGE sample buffer.

Immunostaining and fluorescence microscopy

Cells were grown on segmented cover glass (SCC-002; Matsunami Glass IND.), washed with PBS, and fixed with 4% paraformaldehyde (163-20145; Fujifilm Wako Pure Chemical Corporation) in PBS for 10 min. The cells were permeabilized with 50 µg/ml digitonin in PBS for 5 min and blocked with 3% BSA in PBS for 30 min. Primary antibodies in PBS and 3% BSA were added, and samples were incubated for 1 h at room temperature. Cells were washed five times with PBS, incubated with secondary antibodies in PBS with 3% BSA for 1 h at room temperature, and washed five times with PBS. Rabbit polyclonal antibodies against FIP200 (17250-1-AP; ProteinTech) and mouse monoclonal antibodies against FLAG (F1804; Sigma-Aldrich) were used as primary antibodies for immunostaining. Alexa Fluor 488-conjugated anti-mouse IgG (also cross-adsorbed to rabbit IgG and rat IgG) (A-11029; Thermo Fisher Scientific) and Alexa Fluor 555-conjugated anti-rabbit IgG (also cross-adsorbed to mouse IgG) (A-31572; Thermo Fisher Scientific) were used as secondary antibodies. Cells were observed under a confocal microscope (ECLIPSE Ti2; Nikon).

Halo-LC3 processing assay

Cells were treated with 100 nM tetramethylrhodamine (TMR)-conjugated HaloTag ligand (G8251, Promega) for 15 min. After washing twice with PBS, cells were cultured in the starvation medium for 1 h and lysed as described above. Proteins were separated by SDS-PAGE, and in-gel TMR fluorescence was detected using a fluorescent imaging system (Odyssey M; Li-COR).

Statistical analysis

GraphPad Prism 9 software (GraphPad Software) was used for statistical analyses. The statistical methods are described in each figure legend.

eLife Assessment

This study corroborates recent findings from another group and provides valuable insights into the structure of the autophagy initiation complex, which includes ULK1, ATG13, and FIP200. The authors present solid evidence that supports their claims and addresses one of the key questions in autophagy initiation.

Reviewer #1 (Public review):

In this study, Hama et al. explored the molecular regulatory mechanisms underlying the formation of the ULK1 complex. By employing the AlphaFold structural prediction tool, they showed notable differences in the complex formation mechanisms between ULK1 in mammalian cells and Atg1 in yeast cells. Their findings revealed that in mammalian cells, ULK1, ATG13, and FIP200 form a complex with a stoichiometry of 1:1:2. These predicted interaction regions were validated through both in vivo and in vitro assays, enhancing our understanding of the molecular mechanisms governing ULK1 complex formation in mammalian cells. Importantly, they identified a direct interaction between ULK1 and FIP200, which is crucial for autophagy. However, some aspects of this manuscript require further clarification, validation, and correction by the authors.

Reviewer #2 (Public review):

Summary:

This is important work that helps to uncover how the process of autophagy is initiated - via structural analyses of the initiating ULK1 complex. High-resolution structural details and a mechanistic insight of this complex have been lacking and understanding how it assembles and functions is a major goal of a field that impacts many aspects of cell and disease biology. While we know components of the ULK1 complex are essential for autophagy, how they physically interact is far from clear. The work presented makes use of AlphaFold2 to structurally predict interaction sites between the different subunits of the ULK1 complex (namely ULK1, ATG13, and FIP200). Importantly, the authors go on to experimentally validate that these predicted sites are critical for complex formation by using site-directed mutagenesis and then go on to show that the three-way interaction between these components is necessary to induce autophagy in cells.

Strengths:

The data are very clear. Each binding interface of ATG13 (ATG13 with FIP300/ATG13 with ULK1) is confirmed biochemically with ITC and IP experiments from cells. Likewise, IP experiments with ULK1 and FIP200 also validate interaction domains. A real strength of the work in in their analyses of the consequences of disrupting ATG13's interactions in cells. The authors make CRISPR KI mutations of the binding interface point mutants. This is not a trivial task and is the best approach as everything is monitored under endogenous conditions. Using these cells the authors show that ATG13's ability to interact with both ULK1 and FIP200 is essential for a full autophagy response.

Weaknesses:

I think a main weakness here is the failure to acknowledge and compare results with an earlier preprint that shows essentially the same thing (https://doi.org/10.1101/2023.06.01.543278). Arguably this earlier work is much stronger from a structural point of view as it relies not only on AlphaFold2 but also actual experimental structural determinations (and takes the mechanisms of autophagy activation further by providing evidence for a super complex between the ULK1 and VPS34 complexes). That is not to say that this work is not important, as in the least it independently helps to build a consensus for ULK1 complex structure. Another weakness is that the downstream "functional" consequences of disrupting the ULK1 complex are only minimally addressed. The authors perform a Halotag-LC3 autophagy assay, which essentially monitors the endpoint of the process. There are a lot of steps in between, knowledge of which could help with mechanistic understanding. Not in the least is the kinase activity of ULK1 - how is this altered by disrupting its interactions with ATG13 and/or FIP200?

Reviewer #3 (Public review):

In this study, the authors employed the protein complex structure prediction tool AlphaFold-Multimer to obtain a predicted structure of the protein complex composed of ULK1-ATG13-FIP200 and validated the structure using mutational analysis. This complex plays a central role in the initiation of autophagy in mammals. Previous attempts at resolving its structure have failed to obtain high-resolution structures that can reveal atomic details of the interactions within the complex. The results obtained in this study reveal extensive binary interactions between ULK1 and ATG13, between ULK1 and FIP200, and between ATG13 and FIP200, and pinpoint the critical residues at each interaction interface. Mutating these critical residues led to the loss of binary interactions. Interestingly, the authors showed that the ATG13-ULK1 interaction and the ATG13-FIP200 interaction are partially redundant for maintaining the complex.

The experimental data presented by the authors are of high quality and convincing. However, given the core importance of the AlphaFold-Multimer prediction for this study, I recommend the authors improve the presentation and documentation related to the prediction, including the following:

(1) I suggest the authors consider depositing the predicted structure to a database (e.g. ModelArchive) so that it can be accessed by the readers.

(2) I suggest the authors provide more details on the prediction, including explaining why they chose to use the 1:1:2 stoichiometry for ULK1-ATG13-FIP200 and whether they have tried other stoichiometries, and explaining why they chose to use the specific fragments of the three proteins and whether they have used other fragments.

(3) I suggest the authors present the PAE plot generated by AlphaFold-Multimer in Figure S1. The PAE plot provides valuable information on the prediction.

Author response:

The following is the authors’ response to the original reviews

Public Reviews:

Reviewer #1 (Public review):

In this study, Hama et al. explored the molecular regulatory mechanisms underlying the formation of the ULK1 complex. By employing the AlphaFold structural prediction tool, they showed notable differences in the complex formation mechanisms between ULK1 in mammalian cells and Atg1 in yeast cells. Their findings revealed that in mammalian cells, ULK1, ATG13, and FIP200 form a complex with a stoichiometry of 1:1:2. These predicted interaction regions were validated through both in vivo and in vitro assays, enhancing our understanding of the molecular mechanisms governing ULK1 complex formation in mammalian cells. Importantly, they identified a direct interaction between ULK1 and FIP200, which is crucial for autophagy. However, some aspects of this manuscript require further clarification, validation, and correction by the authors.

Thank you for your thorough evaluation of our manuscript. We have carefully revised the manuscript to address your concerns by performing extra experiments and providing additional clarifications, validations, and corrections as written below.

Reviewer #2 (Public review):

Summary:

This is important work that helps to uncover how the process of autophagy is initiated - via structural analyses of the initiating ULK1 complex. High-resolution structural details and a mechanistic insight of this complex have been lacking and understanding how it assembles and functions is a major goal of a field that impacts many aspects of cell and disease biology. While we know components of the ULK1 complex are essential for autophagy, how they physically interact is far from clear. The work presented makes use of AlphaFold2 to structurally predict interaction sites between the different subunits of the ULK1 complex (namely ULK1, ATG13, and FIP200). Importantly, the authors go on to experimentally validate that these predicted sites are critical for complex formation by using site-directed mutagenesis and then go on to show that the three-way interaction between these components is necessary to induce autophagy in cells.

Strengths:

The data are very clear. Each binding interface of ATG13 (ATG13 with FIP300/ATG13 with ULK1) is confirmed biochemically with ITC and IP experiments from cells. Likewise, IP experiments with ULK1 and FIP200 also validate interaction domains. A real strength of the work in in their analyses of the consequences of disrupting ATG13's interactions in cells. The authors make CRISPR KI mutations of the binding interface point mutants. This is not a trivial task and is the best approach as everything is monitored under endogenous conditions. Using these cells the authors show that ATG13's ability to interact with both ULK1 and FIP200 is essential for a full autophagy response.

Thank you for your thoughtful review and for highlighting the importance of our approach.

Weaknesses:

I think a main weakness here is the failure to acknowledge and compare results with an earlier preprint that shows essentially the same thing (https://doi.org/10.1101/2023.06.01.543278). Arguably this earlier work is much stronger from a structural point of view as it relies not only on AlphaFold2 but also actual experimental structural determinations (and takes the mechanisms of autophagy activation further by providing evidence for a super complex between the ULK1 and VPS34 complexes). That is not to say that this work is not important, as in the least it independently helps to build a consensus for ULK1 complex structure. Another weakness is that the downstream "functional" consequences of disrupting the ULK1 complex are only minimally addressed. The authors perform a Halotag-LC3 autophagy assay, which essentially monitors the endpoint of the process. There are a lot of steps in between, knowledge of which could help with mechanistic understanding. Not in the least is the kinase activity of ULK1 - how is this altered by disrupting its interactions with ATG13 and/or FIP200?

Thank you for this valuable feedback. In response, we performed a detailed structural comparison between the cryo-EM structure reported in the referenced preprint and our AlphaFold-based model. We have summarized both the similarities and differences in newly included figures (revised Figure 2A, B, 3B, S1F) and provided an in-depth discussion in the main text. Furthermore, to address the downstream consequences of ULK1 complex disruption, we have investigated the impact on ULK1 kinase activity, specifically examining how mutations affecting ATG13 or FIP200 interaction alter ULK1’s phosphorylation of a key substrate ATG14. In addition, we analyzed the effect on ATG9 vesicle recruitment. We provide the corresponding data as Figure S3C-E and detailed discussions in the revised manuscript.

Reviewer #3 (Public review):

In this study, the authors employed the protein complex structure prediction tool AlphaFold-Multimer to obtain a predicted structure of the protein complex composed of ULK1-ATG13-FIP200 and validated the structure using mutational analysis. This complex plays a central role in the initiation of autophagy in mammals. Previous attempts at resolving its structure have failed to obtain high-resolution structures that can reveal atomic details of the interactions within the complex. The results obtained in this study reveal extensive binary interactions between ULK1 and ATG13, between ULK1 and FIP200, and between ATG13 and FIP200, and pinpoint the critical residues at each interaction interface. Mutating these critical residues led to the loss of binary interactions. Interestingly, the authors showed that the ATG13-ULK1 interaction and the ATG13-FIP200 interaction are partially redundant for maintaining the complex.

We are grateful for your high evaluation of our work.

The experimental data presented by the authors are of high quality and convincing. However, given the core importance of the AlphaFold-Multimer prediction for this study, I recommend the authors improve the presentation and documentation related to the prediction, including the following:

(1) I suggest the authors consider depositing the predicted structure to a database (e.g. ModelArchive) so that it can be accessed by the readers.

We have deposited the AlphaFold model to ModelArchive with the accession code ma-jz53c, which is indicated in the revised manuscript.

(2) I suggest the authors provide more details on the prediction, including explaining why they chose to use the 1:1:2 stoichiometry for ULK1-ATG13-FIP200 and whether they have tried other stoichiometries, and explaining why they chose to use the specific fragments of the three proteins and whether they have used other fragments.

We appreciate your suggestion. As we noted in the original manuscript, previous studies have shown that the C-terminal region of ULK1 and the C-terminal intrinsically disordered region of ATG13 bind to the N-terminal region of the FIP200 homodimer (Alers, Loffler et al., 2011; Ganley, Lam du et al., 2009; Hieke, Loffler et al., 2015; Hosokawa, Hara et al., 2009; Jung, Jun et al., 2009; Papinski and Kraft, 2016; Wallot-Hieke, Verma et al., 2018). We relied on these findings when determining the specific regions to include in our complex prediction and when selecting a 1:1:2 stoichiometry for ULK1–ATG13–FIP200 which was reported previously (Shi et al., 2020). We also used AlphaFold2 to predict the structures of the full-length ULK1–ATG13 complex and the complex of the FIP200N dimer with full-length ATG13, confirming that there were no issues with our choice of regions (revised Figure S1A-C). In the revised manuscript, we have provided a more detailed explanation of our rationale based on the previous reports and additional AlphaFold predictions.

(3) I suggest the authors present the PAE plot generated by AlphaFold-Multimer in Figure S1. The PAE plot provides valuable information on the prediction.

We provided the PAE plot in the revised Figure S1C.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) In Figure 1D, the labels for the input and IP of ATG13-FLAG should be corrected to ATG13-FLAG FIP3A.

We thank the reviewer for pointing out these labeling mistakes. We revised the labels based on the suggestions.

(2) In the discussion section, the authors should address why ATG13-FLAG ULK1 2A in Fig. 2D leads to a significantly lower expression of ULK1 and provide possible explanations for this observation.

ATG13 and ATG101, both core components of the ULK1 complex, are known to stabilize each other through their mutual interaction. Loss or reduction of one protein typically leads to the destabilization of the other. In this context, ULK1 is similarly stabilized by binding to ATG13. Therefore, ATG13-FLAG ULK2A mutant, which has reduced binding to ULK1, likely loses this stabilizing activity and ULK1 becomes destabilized, resulting in the lower expression levels of ULK1. We added these discussions in the revised manuscript.

(3) In Figure 4B, the authors should explain why Atg13-FLAG KI significantly affects the expression of endogenous ULK1. Could Atg13-FLAG KI be interfering with its binding to ULK1? Experimental evidence should be provided to support this. Additionally, does Atg13-FLAG KI affect autophagy? Wild-type HeLa cells should be included as a control in Figure 4C and 4D to address this question.

Thank you for your constructive suggestion. We found a technical error in the ULK1 blot of Figure 4B. Therefore, we repeated the experiment. The results show that ULK1 expression did not significantly change in the ATG13-FLAG KI. These findings are consistent with Figure S3A. We have replaced Figure 4B with this new data.

We agree that including wild-type HeLa cells as a control is essential to determine whether ATG13-FLAG KI affects autophagy. We performed the same experiments in wild-type HeLa cells and found that ATG13-FLAG KI does not significantly impact autophagic flux. Accordingly, we have replaced Figures 4D and 4E with these new data.

(4) In Figure 3C, the authors used an in vitro GST pulldown assay to detect a direct interaction between ULK1 and FIP200, which was also confirmed in Figure 3E. However, since FLAG-ULK1 FIP2A affects its binding with ATG13 (Fig. 3E), it is possible that ULK1 FIP2A inhibits autophagy by disrupting this interaction. The authors should therefore use an in vitro GST pulldown assay to determine whether GST-ULK1 FIP2A affects its binding with ATG13. Additionally, the authors should investigate whether the interaction between ULK1 and FIP200 in cells requires the involvement of ATG13 by using ATG13 knockout cells to confirm if the ULK1-FIP200 interaction is affected in the absence of ATG13.

Thank you for the valuable suggestion. We examined the effect of the FIP2A mutation on the ULK1–ATG13 interaction using isothermal titration calorimetry (ITC) to obtain quantitative binding data. The results showed that the FIP2A mutation does not markedly alter the affinity between ULK1 and ATG13 (revised Figure S2B), suggesting that FIP2A mainly weakens the ULK1–FIP200 interaction. Regarding experiments in ATG13 knockout cells, ULK1 becomes destabilized in the absence of ATG13, making it technically difficult to assess how the ULK1–FIP200 interaction is affected under those conditions.

Reviewer #2 (Recommendations for the authors):

I feel the manuscript would benefit from a more detailed comparison with the Hurely lab paper - are the structural binding interfaces the same, or are there differences?

We appreciate the suggestion to compare our results more closely with the work from the Hurley lab. We performed a detailed structural comparison between the cryo-EM structure reported in the referenced preprint and our AlphaFold-based model (revised Figure 2A, B, 3B, S1F) and provided an in-depth discussion in the main text.

As mentioned, what happens downstream of disrupting the ULK1 complex? How is ULK1 activity changed, both in vitro and in cells? Does disruption of the ULK1 complex binding sites impair VPS34 activity in cells (for example by looking at PtdIns3P levels/staining)?

Thank you for your insightful comments. We focused on elucidating how disrupting the ULK1 complex leads to impaired autophagy. To assess ULK1 activity, we measured ULK1-dependent phosphorylation of ATG14 at Ser29 (PMID: 27046250; PMID: 27938392). In FIP3A and FU5A knock-in cells, ATG14 phosphorylation was significantly reduced, indicating decreased ULK1 activity (revised Figure S3D, E). This observation is consistent with previous work showing that FIP200 recruits the PI3K complex. Notably, in ATG13 knockout cells, ATG14 phosphorylation became almost undetectable, though the underlying mechanism remains to be fully investigated. Altogether, these data point to reduced ULK1 activity as a key factor explaining the autophagy deficiency observed in FU5A knock-in cells.

We also explored possible downstream mechanisms. One well-established function of ATG13 is to recruit ATG9 vesicles (PMID: 36791199). These vesicles serve as an upstream platform for the PI3K complex, providing the substrate for phosphoinositide generation (PMID: 38342428). To clarify how our mutations impact this step, we starved ATG13-FLAG knock-in cells and observed ATG9 localization. Unexpectedly, even in FU5A knock-in cells where ATG13 is almost completely dissociated from the ULK1 complex, ATG9A still colocalized with FIP200 (revised Figure S3C). These puncta also overlapped with p62, likely because p62 bodies recruit both FIP200 and ATG9 vesicles. Although we suspect that ATG9 recruitment is nonetheless impaired under these conditions, we were unable to definitively demonstrate this experimentally and consider it an important avenue for future study.

Reviewer #3 (Recommendations for the authors):

Here are some additional minor suggestions:

(1) The UBL domains are only mentioned in the abstract but not anywhere else in the manuscript. I suggest the authors add descriptions related to the UBL domains in the Results section.

We thank the reviewer for pointing out the lack of description of UBL domains, which we added in Results in the revised manuscript.

(2) The authors may want to consider adding a diagram in Figure 1A to show the domain organization of the three full-length proteins and the ranges of the three fragments in the predicted structure.

We have added a proposed diagram as Figure 1A.

(3) I suggest the authors consider highlighting in Figure 1A the positions of the binding sites shown in Figure 1B, for example, by adding arrows in Figure 1A.

We have added arrows in the revised Figure 1B (which was Figure 1A in the original submission).

(4) In Figure 1D, "Atg13-FLAG" should be "Atg13-FLAG FIP3A".

We have revised the labeling in Figure 1D.

(5) "the binding of ATG13 and ULK1 to the FIP200 dimer one by one" may need to be re-phrased. "One by one" conveys a meaning of "sequential", which is probably not what the authors meant to say.

We have revised the sentence as “the binding of one molecule each of ATG13 and ULK1 to the FIP200 dimer”.

(6) In "Wide interactions were predicted between the four molecules", I suggest changing "wide" to "extensive".

We have changed “wide” to “extensive” in the revised manuscript.

(7) In "which revealed that the tandem two microtubule-interacting and transport (MIT) domains in Atg1 bind to the tandem two MIT interacting motifs (MIMs) of ATG13", I suggest changing the two occurrences of "tandem two" to "two tandem" or simply "tandem".

We simply used "tandem" in the revised manuscript.

Version 2

Abstract

Abstract

In mammals, autophagosome formation, a central event in autophagy, is initiated by the ULK complex comprising ULK1/2, FIP200, ATG13, and ATG101. However, the structural basis and mechanism underlying the ULK complex assembly have yet to be fully clarified. Here, we predicted the core interactions organizing the ULK complex using AlphaFold, which proposed that the intrinsically disordered region of ATG13 engages the bases of the two UBL domains in the FIP200 dimer via two phenylalanines and also binds the tandem MIT domain of ULK1, thereby yielding the 1:1:2 stoichiometry of the ULK1–ATG13–FIP200 complex. We validated the predicted interactions by point mutations and demonstrated direct triad interactions among ULK1, ATG13, and FIP200 in vitro and in cells, wherein each interaction was additively important for autophagic flux. These results indicate that the ULK1–ATG13– FIP200 triadic interaction is crucial for autophagosome formation and provide a structural basis and insights into the regulation mechanism of autophagy initiation in mammals.

Introduction

Macroautophagy (hereafter autophagy) is an intracellular degradation system delivering cytoplasmic components to the lysosome via autophagosomes (Mizushima and Komatsu, 2011). Autophagy contributes to intracellular quality control by degrading defective proteins and organelles, such as polyubiquitinated proteins and damaged mitochondria. Alternatively, autophagy is induced by starvation as an adaptation mechanism. These physiological roles of autophagy are associated with various diseases, including cancer and neurodegenerative disorders (Mizushima and Levine, 2020). Therefore, understanding and controlling autophagy in clinical contexts is essential.

Autophagosome formation is performed by the evolutionally conserved autophagy-related (ATG) proteins, which form several functional units to act directly on autophagosome formation (Nakatogawa, 2020). Among them, the Atg1/ULK complex is critical in initiating autophagosome formation by organizing the formation site where all the functional units, such as ATG9 vesicles and ATG8 conjugation system, colocalize (Hama, Kurikawa et al., 2023; Kannangara, Poole et al., 2021; Kishi-Itakura, Koyama-Honda et al., 2014; Ren, Nguyen et al., 2023; Suzuki, Kirisako et al., 2001; Yamamoto, Kakuta et al., 2012; Zhou, Liu et al., 2021). In mammals, the ULK complex includes ULK1/2, ATG13, ATG101, and RB1CC1/FIP200 (Mizushima, 2010; Wong, Puente et al., 2013). Structural and functional analysis has made considerable progress in the yeast counterpart, the Atg1 complex, whose core comprises Atg1 (ULK1/2 homolog), Atg13, and Atg17 (remote homolog of FIP200) (Fujioka, Suzuki et al., 2014; Li, Chung et al., 2014; Lin and Hurley, 2016; Noda, 2024; Ragusa, Stanley et al., 2012). Crystallographic analyses have unveiled the detailed interaction mode of Atg13 with Atg1 and Atg17 and the mechanism of the Atg1 complex organization (Fujioka et al., 2014). Moreover, structure-based in vitro and in vivo functional analyses have revealed that the Atg1 complex undergoes liquid-liquid phase separation to organize the autophagosome formation site on the vacuolar membrane to initiate autophagy (Fujioka, Alam et al., 2020). However, structural and functional analyses of the mammalian ULK complex lag behind that of the yeast Atg1 complex, and structural information on the interactions comprising the core of the ULK complex has long been limited to low-resolution electron microscopy (EM) data (Shi, Yokom et al., 2020). A recent preprint reported the cryo-EM structures of the human ULK1 core complex and the supercomplex between the ULK1 core complex and the class III phosphatidylinositol 3-kinase complex 1 at 4.2–6.84 Å resolution, which discovered the unprecedented interaction between these two critical complexes working at autophagy initiation (Chen, Nguyen et al., 2023). Moreover, the local resolution of 3.35 Å enabled them to resolve the ULK1-mediated interactions in the ULK1 core complex. However, not all of the interactions that construct the ULK1 core complex have been perfectly resolved. Recently, calcium transients on the endoplasmic reticulum surface were proposed to trigger the liquid-liquid phase separation of FIP200 to organize the autophagosome formation site, thereby initiating autophagy (Zheng, Chen et al., 2022). However, the lack of structural information makes elucidating the molecular mechanisms underlying these events challenging.

This study predicted the structure of the ULK1–ATG13–FIP200 complex using the AlphaFold2 multimer (Evans, O’Neill et al., 2021), which unveiled the detailed interactions between ATG13 and FIP200, ATG13 and ULK1, and ULK1 and FIP200. In vitro and in vivo mutational analysis confirmed all the predicted interactions, which were demonstrated to be important for autophagy. Furthermore, we found that the FIP200–ATG13 and ULK1–ATG13 interactions partially complement each other for the autophagy initiation. These findings establish the mechanism of ULK complex formation and provide a structural basis for understanding the ULK complex phase separation and regulation mechanism of autophagy in the mammalian context.

Results

Structural prediction of the ULK1–ATG13–FIP200 complex identifies FIP200-interacting residues in ATG13

Previous studies reported that the C-terminal region of ULK1 and the C-terminal intrinsically disordered region of ATG13 bind to the N-terminal region of the FIP200 homodimer and that the stoichiometry of the ULK1–ATG13–FIP200 complex is 1:1:2 (Alers, Loffler et al., 2011; Ganley, Lam du et al., 2009; Hieke, Loffler et al., 2015; Hosokawa, Hara et al., 2009; Jung, Jun et al., 2009; Papinski and Kraft, 2016; Shi et al., 2020; Wallot-Hieke, Verma et al., 2018). However, the structural details underlying the ULK complex assembly, particularly the role of the intrinsically disordered region of ATG13, remain elusive. To understand the mechanism of the ULK complex formation, we predicted the structure of the FIP200 [1–634 amino acids (aa)] dimer complexed with ATG13 (isoform c, 363–517 aa) and ULK1 (801–1,050 aa) (underlined regions in Figure 1A) by AlphaFold2 with the AlphaFold-Multimer option, during which the appropriateness of the regions chosen for our predictions as interaction interfaces was further corroborated by AlphaFold2 models of full-length ATG13 with full-length ULK1 and of full-length ATG13 with the FIP200 (residues 1–634) dimer (Figure S1). Figure 1B depicts the predicted overall structure of the ULK1–ATG13–FIP200 complex, with flexible FIP200 loop regions that received low pLDDT scores omitted for clarity. The overall structure is consistent with the low-resolution cryo-EM model reported previously (Shi et al., 2020), including the C-shape of the FIP200 dimer and the binding of one molecule each of ATG13 and ULK1 to the FIP200 dimer, resulting in the 1:1:2 stoichiometry of the ULK1–ATG13– FIP200 complex. The two FIP200 molecules are named FIP200A and FIP200B, where FIP200A is the one close to ULK1. Extensive interactions were predicted between the four molecules, with a buried surface area of 3,658 Å2 for FIP200A–FIP200B, 2,195 Å2 for FIP200A–ATG13, 1,966 Å2 for FIP200B–ATG13, 2,140 Å2 for ATG13–ULK1, and 1,681 Å2 for FIP200A–ULK1. No interaction was observed between FIP200B and ULK1. One ATG13 molecule binds to the FIP200 dimer through the interaction of residues 365–398 and 394– 482 with FIP200B and FIP200A, respectively, resulting in a 1:2 stoichiometry of the ATG13– FIP200 complex.

Structural basis of the ATG13–FIP200 interaction.

(A) Domain architecture of ULK1, ATG13, and FIP200. Regions used for the AlphaFold2 complex prediction are underlined. (B) Structure of the ULK1–ATG13–FIP200 core complex predicted by AlphaFold2. Flexible loop regions in FIP200 were removed from the Figure for clarity. N and C indicate N- and C-terminal regions, respectively. (C) Close-up view of the interactions between ATG13 and FIP200. The bottom panels represent the surface model of FIP200 with the coloring based on the electrostatic potentials (blue and red indicate positive and negative potentials, respectively). (D) ITC results obtained by titration of MBP-ATG13 (363–517 aa) WT or FIP3A mutant into an FIP200 (1–634 aa) solution. (E) Effect of the ATG13 FIP3A mutation on the FIP200 interaction in vivo. atg13 KO HeLa cells stably expressing FLAG-tagged ATG13 WT or FIP3A were immunoprecipitated with an anti-FLAG antibody and detected with anti-FIP200, anti-ULK1, and anti-FLAG antibodies. (F) Relative amounts of precipitated FIP200 in (E) were calculated. Solid bars indicate the means, and dots indicate the data from three independent experiments. Differences were statistically analyzed using Tukey’s multiple comparisons test.

We compared our AlphaFold2 prediction model with the cryo-EM structures reported as a preprint recently (Chen et al., 2023). For the ULK1–ATG13–FIP200 complex, PDB entries 8SOI, 8SRM, and 8SQZ have global resolutions of 4.2, 4.46, and 5.85 Å, respectively, and employ different ATG13 segments (residues 462–517, 450–517, and 363– 517) (Figure S1F). Despite these differences, all three cryo-EM models contain only the MIM region of ATG13 (residues 462–517). The ULK1 and FIP200 regions used also vary slightly, but are essentially the same as those we used for our AlphaFold2 calculations. In every structure, FIP200 forms a C-shaped dimer whose overall architecture resembles our AlphaFold2 model. While the C-shape opening angle differs among the structures, this variation is consistent with reports that FIP200 can adopt multiple conformations (Shi et al., 2020). The mode of ULK1 binding to FIP200 is likewise fundamentally conserved (a detailed comparison is given below). The main difference among the three structures is their stoichiometry: ULK1:ATG13:FIP200 is 1:1:2 in 8SOI, whereas it is 2:2:2 in the other two entries. In our AlphaFold2 model, ATG13 residues 365–390 bind one FIP200 protomer and residues 442–472 bind the equivalent region of the second protomer, accounting for an ATG13:FIP200 stoichiometry of 1:2. Consequently, a 2:2 stoichiometry is not inconsistent when only the MIM segment of ATG13 is present. Conversely, PDB 8SQZ uses the same ATG13 segment (residues 363–517) as our AlphaFold2 calculations yet displays an ATG13:FIP200 stoichiometry of 2:2. Because this stoichiometry is observed only after the complex has been mixed with PI3KC3-C1 (Chen et al., 2023), some additional regulatory mechanism by PI3KC1-C1 is presumably involved, although its details remain unclear.

Previous studies using hydrogen-deuterium exchange coupled to mass spectrometry identified three regions (M1–M3) in ATG13 responsible for FIP200 binding (Shi et al., 2020). Among these sites, the side-chain of Phe377 in M1 and Phe453 in M3 was deeply inserted into the hydrophobic pockets formed at the base of UBL domains (Figure 1C). Direct interaction between FIP200 (1–634) and Atg13 (363–517) was confirmed by isothermal titration calorimetry (ITC) with a Kd value of 6.6 µM, which was severely attenuated by alanine substitution at these three phenylalanine sites (F377A, F394A, F453A; FIP3A) (Figure 1D, Figure S2A). Furthermore, the ATG13FIP3A mutant expressed in ATG13 KO cells displayed a significantly reduced coprecipitation rate of the endogenous FIP200 (Figure 1E, F). These results confirm that ATG13 and FIP200 interact with each other via the residues predicted by AlphaFold2 in vitro and in vivo.

ULK1 and ATG13 interact via MIT-MIM interaction similar to yeast Atg1–Atg13

Next, we focused on the interaction between ULK1 and ATG13. We previously determined the crystal structure of the yeast Atg1–Atg13 complex, which revealed that the tandem microtubule-interacting and transport (MIT) domains in Atg1 bind to the tandem MIT-interacting motifs (MIMs) of ATG13 (Figure 2A, right) (Fujioka et al., 2014). The predicted structure of the ULK1–ATG13 complex (Figure 2A, left) was quite similar to that of the yeast Atg1–Atg13 complex and to the Cryo-EM structures of the ULK1–ATG13 portion of the ULK1 core complex (Figure 2A, middle). The assembly is stabilized by two sets of the ULK1MIT1– ATG13MIM(C) and ULK1MIT2–ATG13MIM(N) interactions (Figure 2A). Phe470 of ATG13MIM(N) and Phe512 of ATG13MIM(C) were inserted into the hydrophobic pocket of ULK1MIT2 and ULK1MIT1, respectively (Figure 2B). Direct interaction between ULK1 (636–1,050) and Atg13 (363–517) was confirmed by ITC with a Kd value of 0.34 µM, which was severely attenuated by alanine substitution at these two phenylalanine sites (F470A, F512A; ULK2A) (Figure 2C, Figure S2A). In atg13 KO cells, the ULK1 expression level was significantly reduced, suggesting that the lack of ATG13 severely destabilized ULK1. This mirrors the ATG13-mediated stabilization of ATG101, another component of the ULK1 complex (Suzuki, Kaizuka et al., 2015). The exogenous expression of ATG13WT, but not of the ATG13ULK2A mutant, rescued the endogenous ULK1 expression (Figure 2D, E), indicating that the ULK2A mutation in ATG13 attenuates the ATG13–ULK1 interaction in vivo. These observations confirm that, like the yeast Atg1–Atg13 interaction, the ULK1–ATG13 interaction is mediated by MIT–MIM contacts.

Structural basis of the ULK1–ATG13 interaction.

(A) AlphaFold2 model of the ULK1–ATG13 moiety of the ULK1–ATG13–FIP200 core complex in Figure 1 A (left), Cryo-EM structure of the ULK1–ATG13 moiety of the ULK1–ATG13–FIP200 core complex (PDB 8SOI), and crystal structure of the yeast Atg1–Atg13 complex (right, PDB 4P1N). (B) Close-up view of the interactions between ATG13MIM(N) and ULK1MIT2 and between ATG13MIM(C) and ULK1MIT1 (right). (C) ITC results obtained by titration of MBP-ULK1 (636–1050 aa) into a solution of WT or ULK2A mutant of MBP-ATG13 (363–517 aa). Due to weak binding, the KD value for the ULK2A mutant was not accurately determined. (D) Effect of the ATG13-FIP3A mutation on endogenous ULK1 levels in vivo. WT or atg13 KO HeLa cells stably expressing FLAG-tagged ATG13 WT or ULK2A mutant were lysed, and indicated proteins were detected by immunoblotting using anti-FIP200, anti-ULK1, and anti-FLAG antibodies. (E) Relative amounts of ULK1 in (D) were normalized with β-actin and calculated. Solid bars indicate the means, and dots indicate the data from three independent experiments. Differences were statistically analyzed using Tukey’s multiple comparisons test.

Direct interaction between ULK1 and FIP200 is essential for autophagy

Although AlphaFold2 predicted the direct interaction between ULK1 and FIP200 over a reasonably large contact area (1681 Å2) and similar interaction was observed in Cryo-EM structures (Figure 3A, B) (Chen et al., 2023), a previous study reported that ULK1 required ATG13–ATG101 for FIP200 binding in vitro (Shi et al., 2020). ULK1MIT1 and ULK1MIT2 interact directly with FIP200, with the Leu967 of ULK1MIT1 and Phe997 of ULK1MIT2 inserted on the hydrophobic groove of FIP200 (Figure 3A, B). Although we failed to measure the affinity between ULK1 and FIP200 by ITC due to aggregation upon their mixing, an in vitro pull-down assay indicated that ULK1 directly interacted with FIP200, which was significantly attenuated by the FIP2A (L967A, F997A) mutation (Figure 3C, D). Consistently, the ULK1FIP2A mutant lost the interaction with FIP200 while partially retaining the interaction with ATG13 in cells (Figure 3E, F). We confirmed that the ULK1FIP2A mutant retains substantial affinity with ATG13 in vitro by ITC (Figure S2B). These results confirm the direct ULK1–FIP200 interaction in vitro and in vivo. Our results establish the mode of the three independent interactions within the ULK complex, ATG13–FIP200, ATG13–ULK1, and ULK1–FIP200, operating both in vitro and in vivo.

Structural basis of the ULK1–FIP200 interaction.

(A) Structure of the ULK1– FIP200 moiety of the ULK1–ATG13–FIP200 core complex in Figure 1 A. The right panel represents the surface model of FIP200 with coloring based on the electrostatic potentials (blue and red indicate positive and negative potentials, respectively). Dotted squares indicate the regions displayed in (B). (B) Close-up view of the interactions between ULK1MIT1 and FIP200 (top) and between ULK1MIT2 and FIP200 (bottom). Left and right indicate AlphaFold2 and Cryo-EM (PDB 8SOI) models. (C) In vitro pulldown assay between GST-ULK1 (636– 1050 aa) WT or FIP2A mutant with MBP-FIP200 (1–634 aa). (D) Relative amounts of precipitated MBP-FIP200 in (C) were calculated. Solid bars indicate the means, and dots indicate the data from three independent experiments. Differences were statistically analyzed using Tukey’s multiple comparisons test. (E) Effect of the ULK1 FIP2A mutation on the FIP200 interaction in vivo. Ulk1,2 DKO MEFs stably expressing FLAG-tagged ULK1 WT or FIP2A mutant were immunoprecipitated with an anti-FLAG antibody and detected with anti-FIP200, anti-ATG13, and anti-FLAG antibodies. (F) Relative amounts of precipitated FIP200 (left) and ATG13 (right) in (E) were calculated. Solid bars indicate the means, and dots indicate the data from three independent experiments. Differences were statistically analyzed using Tukey’s multiple comparisons test. (G) Halo-LC3 processing assay of ULK1 FIP2A-expressing cells. Ulk1,2 DKO MEFs stably expressing Halo-LC3 and FLAG-tagged ULK1 WT or FIP2A mutant were labeled for 15 min with 100 nm TMR-conjugated Halo ligand and incubated in starvation medium for 1 h. Cell lysates were subjected to in-gel fluorescence detection. (H) Halo processing rate in (G). The band intensity of processed Halo and Halo-LC3 in each cell line was quantified, and the relative cleavage rate was calculated as FLAG-ULK1 WT-expressing cells as 1. Solid bars indicate the means, and dots indicate the data from three independent experiments. Data were statistically analyzed using Tukey’s multiple comparisons test. (I) Colocalization of FLAG-ULK1 WT or FIP2A mutant with FIP200. Ulk1,2 DKO MEFs stably expressing FLAG-tagged ULK1 WT or FIP2A mutant were immunostained with anti-FLAG and anti-FIP200 antibodies. Scale bar, 10 μm.

To investigate the physiological importance of the ULK1–FIP200 interaction, we assessed autophagy flux using the Halo-LC3 processing assay (Rudinskiy, Bergmann et al., 2022; Yim, Yamamoto et al., 2022). Ulk1,2 DKO mouse embryonic fibroblasts (MEFs) expressing FLAG-tagged wild-type ULK1 produce a stronger band of cleaved Halo than cells which do not express ULK1. The intensity of this cleaved Halo band corresponds to the amount of Halo-LC3 delivered to lysosomes in an autophagy-dependent manner. Ulk1,2 DKO MEFs expressing the ULK1FIP2A mutant had partially but significantly reduced processed Halo bands upon starvation compared to ULK1WT-expressing cells, confirming that the direct ULK1–FIP200 interaction is crucial, although partially, for autophagy activity (Figure 3G, H). This partial phenotype would be due to an indirect interaction between ULK1 and FIP200 via ATG13, which was supported by the observation that the ULK1FIP2A mutant colocalized with FIP200 in cells, although less efficiently than ULK1WT (Figure 3I).

Triadic ULK1–ATG13–FIP200 interactions are critical for autophagic flux

We compared the autophagic flux of impaired ATG13–ULK1 and ATG13–FIP200 interactions to confirm the hypothesis that indirect triadic complexes are functional for autophagy. We generated knock-in (KI) cell lines that included ATG13FIP3A, ATG13ULK2A, and ATG13FU5A (F377A, F394A, F453A, F470A, F512A; FU5A) followed by a 3xFLAG-tag in the genomic ATG13 locus of HeLa cells (Figure 4A, B). Since overexpressed ATG13-FLAG is more than 20-fold higher than its endogenous level, the mutation significance can be evaluated at more physiological expression levels using the KI cell lines (Figure S3A, B). The ATG13FIP3A and ATG13ULK2A KI cells displayed partial colocalization with FIP200, whereas the ATG13FU5A KI cells displayed scarce colocalization with FIP200 (Figure 4C). Next, we expressed Halo-LC3 in these KI cells and performed a Halo processing assay. Cleaved Halo was partially reduced in ATG13FIP3A and ATG13ULK2A KI cells compared to ATG13WT cells. In contrast, cleaved Halo in ATG13FU5A KI cells was almost equivalent to that in the knock-out cells (Figure 4D, E). Note that ATG13WT KI cells show ~25% reduction in cleaved Halo than WT HeLa cells, suggesting that FLAG-tag KI only partially affects autophagic flux. These results suggest that ATG13–ULK1 and ATG13–FIP200 bindings complement each other in autophagy function and that ULK1, ATG13, and FIP200 directly bind to each other to organize the robust ULK complex.

ATG13–ULK1 and ATG13–FIP200 interactions complement each other in autophagy function.

(A) Schematic representation of the CRISPR-Cas9-mediated KI strategy of ATG13 mutations with FLAG tag. The C-terminally FLAG-tagged coding sequence after the exon 14 of ATG13 with or without FIP3A, ULK2A, or FU5A mutations were knocked in exon 14 of the Homo sapiens ATG13 locus. As the KI cassette expresses NeoR under the hPGK1 promoter, clones that were successfully knocked in were selected by G418. Cas9-gRNA-targeted sites in the exon 14 of H. sapiens ATG13 locus are displayed in dark blue. The homology arm for KI is presented in magenta, and the ATG13 CDS and mutations in red and cyan, respectively. NeoR is displayed in brown. Scale bar, 0.5 kilobase pair (kb). (B) (B) Immunoblot of ATG13-FLAG KI cell lines. WT, ATG13 KO, and indicated KI HeLa cells were lysed, and indicated proteins were detected by immunoblotting using anti-FIP200, anti-ULK1, and anti-FLAG antibodies. (C) Colocalization of endogenous levels of ATG13-FLAG mutants with FIP200. Indicated KI cell lines were cultured in the starvation medium for 1 h and immunostained with anti-FLAG and anti-FIP200 antibodies. Scale bar, 10 μm. (D) Halo-LC3 processing assay of ATG13-FLAG KI cell lines. WT, ATG13 KO and KI HeLa cell lines were labeled for 15 min with 100 nm TMR-conjugated Halo ligand and incubated in starvation medium for 1 h. Cell lysates were subjected to in-gel fluorescence detection. (E) Halo processing rate in (D). The band intensity of processed Halo and Halo-LC3 in each cell line was quantified, and the relative cleavage rate was calculated as WT HeLa cells as 1. Solid bars indicate the means, and dots indicate the data from three independent experiments. Data were statistically analyzed using Tukey’ s multiple comparisons test. (F) Schematic depiction of the difference between the mammalian ULK complex and the yeast Atg1 complex. Mammalian ATG13 binds to two FIP200s within the same FIP200 dimer, contributing to the stability of one ULK1 complex. Conversely, budding yeast Atg13 binds to two Atg17s within a different Atg17 dimer, allowing for endlessly repeated Atg13-Atg17 interactions. ATG101 in the ULK1 complex and Atg31-29 in the Atg1 complex are omitted for simplicity. ATG13/Atg13 is shown in yellow, ULK1/Atg1 in magenta, and FIP200/Atg17 in green. Black lines represent interactions.

Finally, we investigated how disrupting the ULK complex leads to impaired autophagy. One well-established function of the ULK complex is to recruit ATG9 vesicles (Ren et al., 2023). These vesicles serve as an upstream platform for the PI3KC3-C1, providing the substrate for phosphoinositide generation (Holzer, Martens et al., 2024). To clarify how our mutations impact this step, we starved ATG13 mutant KI cells and observed ATG9A localization. Unexpectedly, even in ATG13FU5A KI cells, where ATG13 is almost completely dissociated from the ULK complex, ATG9A still colocalized with FIP200 (Figure S3C). Since these puncta also overlapped with p62, it is likely that p62 bodies recruit both FIP200 and ATG9 vesicles (Hama et al., 2023). Therefore, it is apparently difficult to assess the autophagosome formation-associated recruitment of the ATG9 vesicles using the present KI cell lines. Another key function of the ULK complex is the ULK1-dependent phosphorylation of ATGs, and phosphorylation of ATG14 at Ser29 is known to be particularly critical (Park, Jung et al., 2016; Wold, Lim et al., 2016). In ATG13FIP3A and ATG13FU5A KI cells, ATG14 phosphorylation was significantly reduced, indicating decreased ULK1 activity (Figure S3D, E). Notably, in ATG13 KO cells, ATG14 phosphorylation became almost undetectable, though the underlying mechanism remains to be fully investigated. Altogether, these data point to reduced ULK1 activity as a key factor explaining the autophagy deficiency observed in ATG13FU5A KI cells.

Discussion

The interaction mechanism of ULK1–ATG13–FIP200, the core of the ULK complex, has been extensively studied (Alers et al., 2011; Chen et al., 2023; Ganley et al., 2009; Hieke et al., 2015; Hosokawa et al., 2009; Jung et al., 2009; Papinski and Kraft, 2016; Shi et al., 2020). However, most of these studies either lacked structural context or, when structural analyses were performed, they were not adequately validated by functional assays in both in vitro and in vivo settings, leaving our understanding of the interaction mechanism still limited. In this study, AlphaFold2-based point mutational analysis provides solid evidence for the direct ULK1–ATG13, ATG13–FIP200, and ULK1–FIP200 bindings in vitro and in vivo. Additionally, the triadic interaction is complementary in the cell.

Regarding the importance of triadic interactions for the molecular function of the ULK complex, we observed their partial significance for ULK1 kinase activity against ATG14, a component of PI3KC3-C1, in our assays (Figure S3D, E). This observation is consistent with recent work showing that FIP200 interacts with the PI3KC3-C1 (Chen et al., 2023). In contrast, we were unable to evaluate autophagosome formation-associated recruitment of the ATG9 vesicles in our ATG13 KI cell lines (Figure S3C). However, given the direct interaction between ATG13 and ATG9A, it is likely that recruitment fails in ATG13FU5A KI cells. In the future, recruitment should be reevaluated under conditions that prevent p62 body– mediated recruitment, such as a penta-receptor KO genetic background.

One major structural difference between the mammalian ULK complex we predicted here and the yeast Atg1 complex is that one ATG13 binds to two FIP200s within the same FIP200 dimer in mammals, whereas one Atg13 binds to two Atg17s in the distinct Atg17 dimers in yeast (Figure 4F). The latter binding mode of Atg13 bridges Atg17 dimers to each other to form a higher-order assemblage (Yamamoto, Fujioka et al., 2016), which is considered to be the Atg1 complex’s phase separation mechanism (Fujioka et al., 2020). Conversely, ATG13 cannot bridge FIP200 dimers to each other and thus cannot induce a higher-order assemblage of the ULK complex, which necessitates another mechanism for phase separation. A recent study reported that calcium transients on the endoplasmic reticulum surface trigger FIP200 phase separation (Zheng et al., 2022); however, the detailed mechanisms are unresolved. Further studies are required to elucidate the mechanisms of phase separation in organizing the autophagosome formation sites in mammals.

In addition to Atg17, budding and fission yeasts have Atg11 as a closer homolog of FIP200 (Kim, Kamada et al., 2001; Li et al., 2014; Pan, Shao et al., 2020; Sun, Li et al., 2013; Yorimitsu and Klionsky, 2005). Atg11 and Atg17 interact with Atg1 and Atg13, respectively, whereas Atg11–Atg13 and Atg17–Atg1 interactions have not been reported. Arabidopsis thaliana, which belongs to the Archaeplastida group, a supergroup distant from Opisthokonta (including yeasts and mammals), also has Atg1, Atg13, and Atg11 (Burki, Roger et al., 2020; Li et al., 2014). The A. thaliana Atg11 interacts with Atg13 but not with Atg1 (Li et al., 2014). The triadic interaction of Atg1/ULK1, ATG13, and Atg11/Atg17/FIP200 has not been reported for any species other than mammals. This triadic interaction is likely beneficial for cells that form the complicated cellular communities that comprise the metazoans. The most likely physiological significance is the fine-tuning of tissue-specific autophagic activity. Consistent with this idea, several splicing variants of ATG13 are present in vertebrates (Alers, Wesselborg et al., 2014; Hieke et al., 2015; Jung et al., 2009). In humans, at least five splicing variants are known, including the isoform 3 lacking the C-terminal ULK1-binding site (MIMs, Figure 2A) (Alers et al., 2014; Jung et al., 2009). The avian Atg13 has seven splicing variants, some lacking parts of the FIP200 interaction site (Alers et al., 2014; Hieke et al., 2015). This study found that a single disruption of the FIP200–ATG13 or ULK1–ATG13 interaction results in a partial reduction of autophagy activity (Figure 4D). In the individual context, the tissue- and organ-specific fine-tuning of the autophagy activity may occur by regulating the expression amount and ratio of splicing variants that lack some of the triad interactions. Generating KI mice with the point mutations identified in this study might help investigate the significance of the triadic binding at the individual level.

Materials and Methods

Structural prediction using AlphaFold2 with the AlphaFold-Multimer mode

Structure was predicted using AlphaFold2 v2.3 installed on a local computer (Sunway Technology Co., Ltd.) (Jumper et al, 2021). The predictions were run using the AlphaFold-Multimer mode (Evans et al, 2021), with five models and a single seed per model, and default multiple sequence alignment generation using the MMSeqs2 server (Mirdita, Schutze et al., 2022; Mirdita, Steinegger et al., 2019). The unrelaxed predicted models were subjected to an Amber relaxation procedure and the relaxed model with the highest confidence based on predicted LDDT scores was selected as the best model and used for figure preparation (Jumper et al, 2021). Structural figures were prepared using PyMOL (http://www.pymol.org/pymol). The model is available in ModelArchive (modelarchive.org) with the accession code ma-jz53c.

Cell culture

HeLa cells and MEFs were cultured in Dulbecco’s Modified Eagle medium (DMEM) (043-30085; Fujifilm Wako Pure Chemical Corporation) supplemented with 10% fetal bovine serum (FBS) (172012; Sigma-Aldrich) in a 5% CO2 incubator at 37°C. Cells were washed twice with phosphate-buffered saline (PBS) and incubated in amino acid-free DMEM (048-33575; Fujifilm Wako Pure Chemical Corporation) without FBS for starvation. atg13 KO HeLa cells (Hama et al., 2023) and Ulk1 Ulk2 double knockout MEFs (Cheong, Lindsten et al., 2011) were described previously.

Plasmids for mammalian cells

Plasmids for stable expression were generated as follows: DNA fragments encoding human ULK1 (NM_003565.4), ATG13 (Hosokawa et al., 2009), and ATG9A (Itakura, Kishi-Itakura et al., 2012) were inserted into the retroviral plasmid pMRX-IP (puromycin-resistant marker) (Kitamura, Koshino et al., 2003; Saitoh, Nakayama et al., 2003) with a 3 × FLAG or 3 x HA epitope tag. Site-directed mutagenesis was used to introduce ATG13FIP3A, ATG13ULK2A, ATG13FU5A, and ULK1FIP2A mutations. pMRXIB-HaloTag7-ratLC3B was described previously (Yim et al., 2022). Plasmids for ATG13 KI were generated as follows: sgRNAs targeting the exon 14 of ATG13 (5′-GGCCTCCCCTCACGATGTCT-3′) were inserted into the BpiI site of PX458 [pSpCas9(BB)-2A-GFP; Addgene #48138]. Donor plasmids for ATG13 KI were generated as follows: 500 base-pair (bp) and 682 bp homology arms were amplified from the genomic ATG13 locus around exon 14 and inserted to flank the KI cassettes (described in Figure 4A) in the donor plasmid (pKnockIn). The coding sequence of 3xFLAG-tagged ATG13 with or without FIP3A, ULK2A, or FU5A mutations was amplified from stable expression plasmids (described above) and inserted just behind the 5’ homology arm.

Stable expression in HeLa cells by retroviral infection

HEK293T cells were transfected with the retroviral plasmid with pCG-gag-pol and pCG-VSV-G (a gift from Dr. T. Yasui, National Institutes of Biomedical Innovation, Health and Nutrition) using FuGENE HD (E2311; Promega) for 4 h in Opti-MEM (31985-070; Gibco). After cell cultivation for 2 days in DMEM, the retrovirus-containing medium was harvested, filtered through a 0.45-µm filter unit (Ultrafree-MC; Millipore), and added to HeLa cells or MEFs with 8 μg/ml polybrene (H9268; Sigma-Aldrich). After 24 h, 2 μg/ml puromycin (P8833; Sigma-Aldrich) or 2.5 μg/ml blasticidin (022-18713; Fujifilm Wako Pure Chemical Corporation) were used to select the stable transformants.

Plasmids for protein preparation

To construct the expression plasmid encoding an N-terminal maltose binding protein (MBP) followed by an HRV3C protease site, the genes were amplified by PCR and cloned into the pET15b vector. The genes were amplified by PCR and cloned into the pET15b-MBP vector for the plasmids encoding N-terminal MBP-tagged ULK1, ATG13, and FIP200. Similarly, for the plasmids encoding N-terminal glutathione S-transferase (GST)-tagged ULK1, the genes were amplified by PCR and cloned into the pGEX6p-1 vector. The NEBuilder HiFi DNA Assembly Master Mix (New England BioLabs) was used to assemble the PCR fragments. PCR-mediated site-directed mutagenesis was used to introduce the mutations that led to the specified amino acid substitutions. All constructs were sequenced to confirm their identities.

Protein expression and purification

E. coli BL21 (DE3) cells were used to express all recombinant proteins. After cell lysis, MBP-FIP200 (1–634) was purified by affinity chromatography using Amylose Resin High Flow (New England Biolabs). Next, MBP was cleaved with the human rhinovirus 3C protease. The eluates were then desalted with 20 mM Tris-HCl, pH 8.0, and 150 mM NaCl utilizing a Bio-Scale Mini Bio-Gel P-6 desalting column (Bio-Rad Laboratories). Subsequently, the cleaved MBP was removed by reappliying FIP200 (1–634) to an Amylose Resin High Flow column. For the pull-down assay, MBP-FIP200 (1–634) eluted from the amylose resin continued to be purified on a HiLoad 26/60 Superdex 200 PG column eluted with 20 mM Tris-HCl, pH 8.0, and 150 mM NaCl.

Affinity chromatography with Amylose Resin High Flow (New England Biolabs) was used to purify the MBP-Atg13 (363–517) and MBP-ULK1 (636–1,050) proteins. They were further purified on a HiLoad 26/60 Superdex 200 PG column and eluted with a buffer containing 20 mM Tris-HCl, pH 8.0, and 150 mM NaCl. Similarly, GST-ULK1 (636–1050) underwent initial purification using a GST-accept resin (Nacalai Tesque, 09277-14). This step was followed by further purification on a HiLoad 26/60 Superdex 200 PG column using the same elution buffer as above.

In vitro pull-down assay

Purified proteins were incubated with GST-accept beads (Nacalai Tesque) at 4°C for 30 min. The beads were washed three times with PBS, and proteins were eluted with 10 mM glutathione in 50 mM Tris-HCl (pH 8.0). SDS-PAGE was used to separate the samples, and protein bands were detected by One Step CBB (BIO CRAFT).

Isothermal titration calorimetry

The binding of FIP200 (1–634) to MBP-ATG13 (363–517) and MBP-ATG13 (363–517) to MBP-ULK1 (636–1,050) was measured by ITC, with a MICROCAL PEAQ-ITC calorimeter (Malvern) and stirring at 750 rpm at 25°C. All proteins were prepared in a solution of 20 mM Tris-HCl, pH 8.0, and 150 mM NaCl. The titration of MBP-ATG13 (363–517) with FIP200 (1– 634) involved 18 injections of 2 μl of the MBP-ATG13 (363–517) solution (229 μM) at 150 s intervals into a sample cell containing 300 μl of FIP200 (1–634) (4 μM). The titration of MBP-ULK1 (636–1,050) with MBP-ATG13 (363–517) involved 18 injections of 2 μl of the MBP-ULK1 (636–1,050) WT at 232 μM or FIP2A mutant at 172 μM at 150 s intervals into a sample cell containing 300 μl of MBP-ATG13 (363–517) at 22.9 μM or 20.0 μM, respectively. The isotherm was integrated and fitted with the one-side-binding model of the Malvern MicroCal PEAQ-ITC analysis software. The error of each parameter indicates the fitting error.

Generation of KI cell lines

HeLa cells were transfected with the PX458-based plasmid expressing sgRNA for the ATG13 gene and donor plasmids using FuGENE HD (E2311; Promega). One day after transfection, the KI cells were selected with 1 mg/mL G418 for 14 days. After single clones were isolated, clones positive for the FLAG tag and negative for wild-type ATG13 were screened by immunoblotting.

Preparation of whole cell lysates and immunoblotting

HeLa cells were lysed with 0.2% n-octyl-β-D-dodecyl maltoside in lysis buffer [50 mM Tris-HCl (pH 7.5), 150 mM NaCl, 1 mM EDTA, EDTA-free protease inhibitor cocktail (04080; Nacalai Tesque)] for 15 min on ice. After centrifugation at 20,000 × g for 15 min at 4°C, the supernatants were mixed with 20% volumes of 6 × SDS-PAGE sample buffer. The samples were subjected to SDS-PAGE and transferred to Immobilon-P PVDF membranes (IPVH00010; EMD Millipore). The primary antibodies used for immunoblotting were rabbit polyclonal antibodies against FIP200 (17250-1-AP; ProteinTech), ATG13 (13273; Cell Signaling), ULK1 (8054; Cell Signaling), ATG14 p-Ser29 (#92340; Cell Signaling), ATG14 (24412-1-AP; ProteinTech), and mouse monoclonal antibody against FLAG (F1804; Sigma-Aldrich) and β-actin (A2228; Sigma-Aldrich). Horseradish peroxidase (HRP)-conjugated anti-rabbit IgG (111-035-144; Jackson ImmunoResearch) was used as a secondary antibody. The HRP-conjugated mouse monoclonal antibody against DDDDK-tag (M185-7; MBL) was used for FLAG tag detection. Immobilon Western Chemiluminescent HRP Substrate (P90715; EMD Millipore) was used to visualize the signals detected by an image analyzer (ImageQuant LAS 4000; Cytiva). Fiji software (ImageJ; National Institutes of Health) was used to adjust contrast and brightness and quantify the signals (Schindelin, Arganda-Carreras et al., 2012).

Immunoprecipitation

Cells were lysed with 1% 3-[(3-cholamidopropyl)-dimethylammonium]-1-propanesulfonate (CHAPS) in lysis buffer and centrifuged at 20,000 g for 15 min. The supernatants were incubated with anti-FLAG M2 affinity gel (A2220; Sigma-Aldrich) for 1 h at 4°C with gentle rotation. The beads were washed three times in washing buffer (20 mM Tris-HCl, pH 8.0, 150 mM NaCl, 1 mM EDTA, and 0.5% CHAPS), and the proteins were eluted with the SDS-PAGE sample buffer.

Immunostaining and fluorescence microscopy

Cells were grown on segmented cover glass (SCC-002; Matsunami Glass IND.), washed with PBS, and fixed with 4% paraformaldehyde (163-20145; Fujifilm Wako Pure Chemical Corporation) in PBS for 10 min. The cells were permeabilized with 50 µg/ml digitonin in PBS for 5 min and blocked with 3% BSA in PBS for 30 min. Primary antibodies in PBS and 3% BSA were added, and samples were incubated for 1 h at room temperature. Cells were washed five times with PBS, incubated with secondary antibodies in PBS with 3% BSA for 1 h at room temperature, and washed five times with PBS. Rabbit polyclonal antibodies against FIP200 (17250-1-AP; ProteinTech) and mouse monoclonal antibodies against FLAG (F1804; Sigma-Aldrich) HA (M180-3, MBL) and guinea pig antibody against p62 (GP62-C; PROGEN) were used as primary antibodies for immunostaining. Alexa Fluor 488-conjugated anti-mouse IgG (also cross-adsorbed to rabbit IgG and rat IgG) (A-11029; Thermo Fisher Scientific), Alexa Fluor 555-conjugated anti-rabbit IgG (also cross-adsorbed to mouse IgG) (A-31572; Thermo Fisher Scientific) and Alexa Fluor 647-conjugated anti-guinea pig (A21450; Thermo Fisher Scientific) were used as secondary antibodies. Cells were observed under a confocal microscope (ECLIPSE Ti2; Nikon).

Halo-LC3 processing assay

Cells were treated with 100 nM tetramethylrhodamine (TMR)-conjugated HaloTag ligand (G8251, Promega) for 15 min. After washing twice with PBS, cells were cultured in the starvation medium for 1 h and lysed as described above. Proteins were separated by SDS-PAGE, and in-gel TMR fluorescence was detected using a fluorescent imaging system (Odyssey M; Li-COR).

Statistical analysis

GraphPad Prism 9 software (GraphPad Software) was used for statistical analyses. The statistical methods are described in each figure legend.